A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway using an accessible web application

### 2018-03-10 manuscript/manuscript.Rmd
added a code chunk in sister pair diversification
last updates before sending to google drive

# List of working titles  
* A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway  
* *Add your suggestions*

@@

The first genome sequence of *Arabidopsis thaliana* facilitated rapid advancement of plant biology through molecular genetics. Since this initial genome, massive scaling of sequencing technology has allowed a pioneering group to survey the global genomic variation in natural *Arabidopsis thaliana* populations. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits. Beyond these genome-wide studies for gene- and variant- discovery this valuable dataset has seen little use. This natural genomic variation also provides a set of feasible functional variation at the gene, gene family, and gene network level that may provide insight into the function and evolution of genes, families, and networks. We present a web application and R-package through which plant molecular biologists with little to no bioinformatics experience can make use of this rich dataset of genetic variation to formulate hypotheses as to the sequence/function/phenotype relationships determined by the gene, family, or network of their interest. This candidate gene/family tool also allows family-wise alignments of variants to facilitate de novo functional domain identification and evolutionary analysis, easy access to lists of missense polymorphisms to facilitate biochemical assays of variant effects, as well as lists of accessions containing missense (or any type of) polymorphisms to facilitate genetic segregation analysis for deterimining the phenotypic effects of variants and the study of gene network evolution. Here, we have utilized this tool to analyze the genetic variation in the nuclear auxin signaling pathway, and formulate hypotheses regarding the functional implications of this variation and the evolution of the genes and gene families in this pathway.

# Related work  
What and how have we, as a field, learned from natural variation?  
[@hollister_genomic_2014; @bielawski_detecting_2013; @kryazhimskiy_population_2008; @schmid_multilocus_2005; @nieduszynski_sequence_2011; @melamed_combining_2015; @weigel_population_2015; @schmitz_patterns_2013; @weigel_natural_2005; @hardtke_case_2010; @koornneef_naturally_2004; @mitchell-olds_genetic_2006; @joly-lopez_developing_2016; @wright_insights_2017a; @park_distributions_2017; @_135_2016a]

@@

where $N_{ik}$ is the nucelotide (A, T, C or G) at position k on the ith sequence of the population.
$L$ is the lenght of the sequence (the same for all sequences in the population, 
assumes only SNPs exist, no indels). n is the total number of sequences in the population.

from this form we can re-arange summations to the form below:

@@

here $\pi_k$ can be thought of as the site-wise nucleotide diversity at postion k, and is equal to 
the nucleotide diversity of a sequence of lenght 1 at location k. we can calculate $\pi_k$ for each
site, then average those over the sequence lenght to calculate $\pi$, the nucleotide diversity 
of the sequence.

@@

The formula for $\pi_k$ above requires comparing every sequence to everyother sequence at location k,
however, we know there are only a few variant forms at each individual location.

so we can revert back to using Nei's original forula, modifying it slightly,
replacing $x_i$ with $\frac{n_i}{n}$, $n_i$ being the number of sequences in the population of the form $N_i$ at location k: subscripts i and k are summed only over unique forms [need to clarify this]

@@

this simplified form for $\pi_k$ is used by the app, because the counts of unique 
variants at a single nucleotide location can easily be sumaraized in R.

@@

Auxin acts by binding to receptors that in turn target co-repressors for degradation. Auxin receptors (six in the model plant Arabidopsis thaliana) evolved through gene duplication and diversification early in the history of vascular plants [@Parry_Complex_2009]. The rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor [@Havens_Synthetic_2012], and that this rate sets the pace of development [@Guseman_Auxin-induced_2015].

@@

Examining the natural sequence variation across the AFB revealed that *TIR1* and *AFB1* both had very low nonsynonymous diversity, hinting at their likely functional importance. *AFB3* and *AFB4* had higher nonsynonymous diversity, while their sisters genes, *AFB2* and *AFB5* are more conserved. This matches our current understanding of *AFB3* and *AFB4* as being minor players in the auxin signaling pathway, but brings in to question the inconclusive role of *AFB1* in auxin signaling. Although most important functional regions do not contain variants in AFB1, there is a low diversity mutant, only found in the Can-1 accession, which is in the highly conserved TIR1/AFB oligomerization domain. Because variants eliciting significant changes in function often are not selected for, the fact that this variant is extremely low diversity may suggest that it is associated with a strong auxin signaling phenotype. Characterization of this accession may identify potential *AFB1* phenotypes and may also test the hypothesis that AFB1 functions through oligomerization with other TIR1/AFBs.
  The AFB4 and AFB5 receptors have an N-terminal extension prior to the F-box domains. This extesion had very high nonsynonymous diversity(Fig. 4), suggesting that this extension does not play an important functional role in these proteins.

Figure 4: Alignment of A. thaliana TIR1/AFB F-box domains showing variants. Cyan marks the Fbox binding domain. Yellow, green, and magenta, mark missense and synonymous, synonymous only, and missense only variant locations respectively. Yellow squared E’s show Cul1-interacting residues, differing in AFB1.

@@

###Introduction
The Aux/IAA proteins (IAAs) mediate both auxin perception and transcriptional co-repression of auxin responsive genes. Auxin perception is carried out along with TIR1/AFB proteins, which together with IAAs act as a coreceptor to bind and perceive intracellular auxin. Interaction of IAAs with auxin and TIR1/AFBs is conferred by the highly conserved 13 amino acid degron motif [@tan_mechanism_2007]. Formation of this auxin coreceptor complex triggers ubiquitination and subsequent degradation of IAAs. This degradation in turn relieves their repression upon ARF transcription factors.Stabilizing mutations within the degron can lead to dominant phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. The Arabidopsis Aux/IAA gene family has 29 members, most of which possess a canonical degron. Notable exceptions are the auxin-insensitive IAA20 and IAA30 that lack a degron sequence entirely, while the closely related IAA31 retains a semi-functional degron sequence with a very slow degradation rate [@dreher_arabidopsis_2006; @havens_synthetic_2012]. Additionally, IAA degradation rate is influenced by rate motifs found outside the degron, such as the KR motif [@moss_rate_2015; @dreher_arabidopsis_2006]. IAA-mediated repression of ARFs depends upon both an EAR domain, N-terminal of the degron, that recruits co-repressors of the TOPLESS (TPL) family, and a Phox and Bem1 (PB1) domain that enables interaction via heteroligomerization with ARF PB1 domains [@nanao_structural_2014; @kagale_ear_2011].

@@

###Sister Pair Diversification
Previous research has ~~pointed to~~ *found evidence of* genetic redundancy within closely related pairs or groups of IAA proteins [@overvoorde_functional_2005]. For example, the iaa8-1, iaa9-1 double mutant and the iaa5-1, iaa6-1, iaa19-1 triple mutant have wild-type phenotypes [@overvoorde_functional_2005]. Despite this, most *Arabidopsis* IAA sister pairs have one member with high nonsynonymous diversity while the other sister is more conserved (Figure IAA phylogeny heat map). For example, despite having only minor differences in sequence (61.4% sequence identity), the IAA6/19 sister pair has significant differences in expression patterns, protein abundances and functions, suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017]. In an interspecific comparison between *A. thaliana* abd *A. lyrata*, IAA19 is more conserved, while IAA6 showed regions of increased sequence divergence upstream of the degron and in the N-terminal half of the PB1 domain [@winkler_variation_2017]. Consistent with these results, our analysis of intraspecific natural variation revealed higher conservation for IAA19 (${\pi_N/\pi_S}$ = 0.55) compared to IAA6 (${\pi_N/\pi_S}$ = 2.3) (Figure IAA phylogeny heat map), and also detected high diversity within the same regions of IAA6 as seen in Winkler et al. (Supp Fig IAA6 diversity plot).

Within the Group B IAA genes, there are four sister pairs--IAA10/IAA11, IAA12/IAA13, IAA18/IAA26, and IAA20/IAA30 (Fig X) [@remington_contrasting_2004]. Interestingly, among these sister pairs only IAA12 and 13 have a noticeable difference in nonsynonymous diversity. The Arabidopsis EFP browser shows that IAA10 and 11 have almost identical expression patterns, with the exception of IAA10 being more highly expressed in the developing seed, and IAA11 being more highly expressed in root tissue during lateral root development. In our analysis, IAA10 and IAA11 both showed functional conservation (${\pi_N/\pi_S}$ of 0.80 and 0.67 respectively) in support of IAA10 and IAA11 playing redundant roles. Accessions and associated variants identified here provide a potentially expedited route to elucidating the extent of redundancy between family members such as IAA10 and 11.

###Identification of new Loss-of-Function/Gain-of-Function mutations
Accessions with functional variants can be potentially used to investigate the extent of redundancy/specialization within sister pairs. For example, it has been shown that IAA10 has a missense mutation in the EAR domain, which converts the second leucine in the LxLxL motif to a serine (L49S), a mutation likely to reduce the ability of the IAA to recruit the TOPLESS co-repressor [@lee_biological_2016]. This mutation was found in three accessions from the southern UK. A second potentially stronger mutation, represented in one accession from northeastern Spain is a ~~missense~~ *frameshift?* resulting in an early stop codon in the PB1 domain (need Supp. Fig?). This likely prevents IAA10 from interacting with an ARF. ~~These alleles would be strong candidates for loss-of-function mutants, allowing researchers to discover the function of IAA10. In particular, it would make possible a straightforward path for creating an iaa10/11 double mutant using CRISPR/Cas9 directed mutagenesis, allowing for examination of the extent of redundancy between these family members.~~ *I don't think we can say this. There are SALK lines in the 5' UTR and first intron of IAA10 and IAA11.*

Several examples of possible gain-of-function degron missense variants were found in IAA2, IAA3, IAA6, IAA7, IAA8, IAA13, IAA15, IAA18, IAA19, IAA27, IAA28, IAA29. Most degron variants were only found in one or two accessions (Supplemental Table or Map?). ~Interestingly, almost all of the IAAs found to have accessions with degron missense mutations have a closely related sister, which may indicate a redundant function within the sister-pairs.~ *But we know that iaa6 iaa19 iaa5 triple doesn't have a phenotype.* IAA6 has two potentially dominant missense mutations found within the degron (Supp Fig. alignment), one previously believed to lead to the shy1-1 phenotype [@kim_two_1996]. Interestingly, this variant is widely represented in 31 accessions across europe (Supplemental Map Fig). This finding supports the proposition that iaa6[C78R] may not be the mutation causing the shy1-1 phenotype but rather represents natural variation in the degron sequence ([note from Mike Prigge in TAIR](https://www.arabidopsis.org/servlets/TairObject?type=notepad&id=11247)). These accessions could be used to confirm whether iaa6[C78R] is responsible for the SHY1-1D phenotype. Additionally, an accession was identified with the IAA6-V77M variant, also in the degron, that may be useful to further explore iaa6 gain-of-function phenotypes.

@@

Members of the IAA17/7/14/16 group have a second EAR motif (LxLxL) with no observed missense mutations in IAA7, 16, or 17 (Supp Fig IAA7/14/16/17 alignment). This observation supports the experimental evidence for IAA7 that the second EAR motif is functional: it can interact with TPR1 corepressor and has a minor repressive role in auxin-related developmental phenotypes [lee_biological_2016]. For IAA14, we did observe two missense mutations that may be of interest to probe functionality of the IAA14 EAR motif: one in the central lysine (L41F) that is represented in only one accession, and the other in the second “x” (K40N) that is found in 12 accessions around Copenhagen (Supp Fig. IAA14 map).

@@

  The Auxin signaling pathway utilizes the TOPLESS (TPL) family of Gro/TLE/TUP1 type co-repressor proteins to maintain auxin responsive genes in a transcriptionally-repressed state in the absence of auxin (cite). In Arabidopsis the TPL family includes TOPLESS and four TPL-related proteins (TPR1-4), which are comprised of three structural domains: an N-terminal TPL domain and two beta-propellar domains. TPL is recruited to the AUX/IAA proteins through interaction with a conserved protein motif known as an Ethylene-responsive element binding factor-associated amphiphilic repression (EAR) domain (cite). Canonical EAR domains have the amino acid sequence LxLxL, as is found in domain I of the AUX/IAAs (figure xA, cite). TPL co-repressors bind EAR domains through direct interaction with the TPL N-terminal C-terminal to LisH (CTLH) domain (citations of pre-structure founding papers/reviews). Recent structural analyses of the TPL N-terminal domain have highlighted the precise interaction interface between TPL and AUX/IAA EAR domains, as well as the TPL-TPL dimerization and tetramerization motifs (Avarelilo et al 2017, Ke et al, 2015), and even the residues required for higher-order multimers of TPL tetramers (Ma et al, 2017). The two tandem beta propellar domains of TPL/TPRs have as yet not been crystallized, however numerous high resolution crystal structures of beta propellar domains allow for structural prediction of the TPL c-terminal beta propellar domains and the locations of likely interaction faces (Figure, X).

  The TOPLESS co-repressor family generally exhibits a high level of sequence conservation at the amino acid sequence level across the Arabidopsis accessions with all ${\pi_N/\pi_S}$ values below 1 (Table X). The closely related TPL and TPR1 have the highest ${\pi_N/\pi_S}$ values (TPL-0.425, TPR1-0.739), suggesting that these these two related genes tolerate a higher degree of sequence diversity compared to TPR2-4. The TPL/TPR family demonstrates a high level of sequence conservation across the N-terminal TPL domain, with the lowest of observed ${\pi_N/\pi_S}$ values (Figure X). Indeed, all non-synonymous mutations observed in this region are either in linker domains between alpha helices, or are highly conservative mutations within helices (i.e. Valine to Isoleucine), which would be predicted to exhibit little effect on TPL folding and activity.

  The increased N-terminal conservation underscores the importance of the N-terminal TPL domain. For example, the initial tpl-1 mutation (N176H) in the ninth helix is a dominant gain-of-function allele (Long et al, 2002, PMID12050130), which is capable of binding wild-type TPL protein and inducing protein agreggation (Ma, 2017, pmid28630893). It is therefore understandable that very little diversity is observed in the ninth alpha helix of TPL/TPRs as mutations in this domain could act in a dominant negative fashion, rendering any de novo mutation in this region embryo lethal. However, the entire N-terminal domain exhibits very low rates of nucleotide diversity, suggesting that each domain in this region is critical at the population level. Indeed, the crystal structure of TPL has revealed that the TPL domain is responsible for the multiple roles of TPL/TPR proteins: The LisH domain and CRA domain coordinate TPL dimerization and tetramerization respectively (26601214, 28698367), the CTLH domain is critical for interaction with DNA-binding transcription factors (26601214), and helix 9 within the CRA domain may be critical for the formation of higher-order TPL complexes (multimers of TPL tetramers, 28630893).

  We then compared the frequency of SNPs that cause nonsynonymous mutations across the TPL domains. Compared to the TPL domain, both the proline rich and WD-40 domains exhibit a higher rate of missense mutations accross both of the WD-40 repeat regions, here labeled WD40-1 and WD40-2 respectively. We utilized the Phyre2.0 modeling software to generate a rough model of the TPL/TPR WD-40 repeats stricly for the purpose of mapping non-synonymous mutations onto a visible protein chassis (Figure X). Here we can see both faces of each predicted beta-propellar domains, which are connected by a short linker region. Through utilization of protein modelling, it becomes clear that we observe conservation of amino acid sequence at the center of the beta-propellar domains. These observations could be used as the basis for forward mutagensis strategies to test whether this region of TPL functions similarly to its homolog TLE, which binds its target genes EH1 and the C-terminal WRPW/Y motif (Hairy/Hes/Runx) in the center of the beta-propeller domain (pmid 16762837).

  It is interesting to note that despite the presence of 5 TPL/TPR genes in the genome, which are considered to be redundant (pmid 16763149), very few SNPs are observed that are predicted to result in loss of function for any one TPL/TPR gene. In fact, only one deletion was observed (665delG) that creates a frameshift mutation in TPR4 in a single accession (Aiell-4 in Italy). This mutation occurs in codon 222, just downstream of the TPL N-terminal domain, and would generate an early stop codon 11 codons downstream of the deletion site. It is unclear whether this N-terminal fragment would be expressed at any appreciable level, as it is possible that the generation of an early stop codon would trigger nonsense-mediated decay. However, if expressed, it would be fascinating to examine whether this naturally occuring N-terminal truncation retains activity as a co-repressor. Given the observation that the TPL N-terminal domain is sufficient to bind histones (28630893), and repress transcription via synthetic assays in yeast (24979769), it is possible that this represents a semi-functional TPL variant. There has been no evidence for priotization of individual TPL/TPR family members in the auxin response pathway, but experiments modulating the gene copy number of TPR1 revealed the importance of maintaing the proper dosage of TPL/TPR in order to accurately trigger the salicyclic acid innate immune response (20647385). These results, taken together with the absence of TPL/TPR loss-of-function alleles in the Arabidopsis accessions suggest that perhaps the entire family is being conserved at the nucleotide level in the population to meet a gene dosage requirement, with little opportunity for neo- or sub-functionalization.

@@

Auxin response is mediated by the auxin responsive transcription factors, or ARFs. There are 23 ARFs in Arabidopsis that are divided into three phylogenetic classes. Class A ARFs, ARF5, ARF6, ARF7, ARF8, and ARF19, activate transcription. These ARFs have a Q-rich region in the middle of the protein that may mediate activation (Guilefoyle, 2007). It has recently been shown that the middle region of ARF5 interacts with the SWI/SNF chromatin remodeling ATPases BRAMA and SPLAYED, possibly to reduce nucleosome occupancy and allow for the recruitment of transcription machinery (Wu, 2016). Additionally ARF7 can interact with Mediator subunits, directly tethering transcriptional proteins to its binding regions (Ito, 2016). Class B and C ARFs are historically categorized as repressor ARFs, though the mechanism through which they confer repression has not been identified. Their middle regions tend to be proline- and serine-rich.

>>>>

# List of working titles  
* A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway  
* **Add your suggestions**

@@

The first genome sequence of *Arabidopsis thaliana* facilitated rapid advancement of plant biology through molecular genetics. Since this initial genome, massive scaling of sequencing technology has allowed a pioneering group to survey the global genomic variation in natural *Arabidopsis thaliana* populations. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits. Outside of genome-wide studies this valuable dataset has seen little use. Natural genomic variation also provides a set of feasible functional variation at the gene, gene family and gene network level that may provide insight into the function and evolution of genes, families and networks. Here, we present a web application and R-package through which plant molecular biologists with little to no bioinformatics experience can make use of this rich dataset of genetic variation to formulate hypotheses as to the sequence/function/phenotype relationships determined by the gene, family, or network of their interest. We demonstrate the utility of this tool through comprehensive analysis and identification of potential functional variation in the nuclear auxin signaling pathway.

This application allows easy access to 
1. lists of missense polymorphisms to facilitate biochemical assays of variant effects [@starita_variant_2017],
2. family-wise alignments of variants to facilitate de novo functional domain identification [@melamed_combining_2015], 
3. lists of accessions containing missense (or any type of) polymorphisms to facilitate segregation analysis and measurment of the distribution of variant effects on phenotype [@park_distributions_2017], 
4. as well as the study of gene and network evolution [@delker_natural_2010; @kliebenstein_role_2008].

Here, we have utilized this tool to analyze the genetic variation in the nuclear auxin signaling pathway and formulate hypotheses regarding the functional implications of this variation and the evolution of the genes and gene families in this pathway. Our analysis also provided further confirmation of much of the existing knowledge of these genes demonstrating the validity of this approach.

@@

where $N_{ik}$ is the nucleotide (A, T, C or G) at position k on the ith sequence of the population.
$L$ is the length of the sequence (the same for all sequences in the population, 
assumes only SNPs exist, no indels). n is the total number of sequences in the population.

from this form we can re-arrange summations to the form below:

@@

here $\pi_k$ can be thought of as the site-wise nucleotide diversity at position k, and is equal to 
the nucleotide diversity of a sequence of length 1 at location k. we can calculate $\pi_k$ for each
site, then average those over the sequence length to calculate $\pi$, the nucleotide diversity 
of the sequence.

@@

The formula for $\pi_k$ above requires comparing every sequence to every other sequence at location k,
however, we know there are only a few variant forms at each individual location.

so we can revert back to using Nei's original formula, modifying it slightly,
replacing $x_i$ with $\frac{n_i}{n}$, $n_i$ being the number of sequences in the population of the form $N_i$ at location k: subscripts i and k are summed only over unique forms [need to clarify this]

@@

this simplified form for $\pi_k$ is used by the app, because the counts of unique 
variants at a single nucleotide location can easily be summarized in R.

@@

Auxin acts by binding to receptors that in turn target co-corepressors for degradation. Auxin receptors (six in the model plant *Arabidopsis thaliana*) evolved through gene duplication and diversification early in the history of vascular plants [@Parry_Complex_2009]. The rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor [@Havens_Synthetic_2012], and that this rate sets the pace of development [@Guseman_Auxin-induced_2015].

@@

Examining the natural sequence variation across the AFB revealed that *TIR1* and *AFB1* both had very low nonsynonymous diversity, hinting at their likely functional importance. *AFB3* and *AFB4* had higher nonsynonymous diversity, while their sisters genes, *AFB2* and *AFB5* are more conserved. This matches our current understanding of *AFB3* and *AFB4* as being minor players in the auxin signaling pathway, but brings in to question the inconclusive role of *AFB1* in auxin signaling. Although most important functional regions do not contain variants in AFB1, there is a low diversity mutant, only found in the Can-1 accession, which is in the highly conserved TIR1/AFB oligomerization domain. Because variants eliciting significant changes in function often are not selected for, the fact that this variant is extremely low diversity may suggest that it is associated with a strong auxin signaling phenotype. Characterization of this accession may identify potential *AFB1* phenotypes and may also test the hypothesis that AFB1 functions through oligomerization with other TIR1/AFBs.
  The AFB4 and AFB5 receptors have an N-terminal extension prior to the F-box domains. This extension had very high nonsynonymous diversity(Fig. 4), suggesting that this extension does not play an important functional role in these proteins.

Figure 4: Alignment of A. thaliana TIR1/AFB F-box domains showing variants. Cyan marks the F-box binding domain. Yellow, green, and magenta, mark missense and synonymous, synonymous only, and missense only variant locations respectively. Yellow squared E’s show Cul1-interacting residues, differing in AFB1.

@@

The Aux/IAA proteins (IAAs) mediate both auxin perception and transcriptional co-repression of auxin responsive genes. Auxin perception is carried out along with TIR1/AFB proteins, which together with IAAs act as a co-receptor to bind and perceive intracellular auxin. Interaction of IAAs with auxin and TIR1/AFBs is conferred by the highly conserved 13 amino acid degron motif [@tan_mechanism_2007]. Formation of this auxin co-receptor complex triggers ubiquitination and subsequent degradation of IAAs. This degradation in turn relieves their repression upon ARF transcription factors.Stabilizing mutations within the degron can lead to dominant phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. The *Arabidopsis thaliana* Aux/IAA gene family has 29 members, most of which possess a canonical degron. Notable exceptions are the auxin-insensitive IAA20 and IAA30 that lack a degron sequence entirely, while the closely related IAA31 retains a semi-functional degron sequence with a very slow degradation rate [@dreher_arabidopsis_2006; @havens_synthetic_2012]. Additionally, IAA degradation rate is influenced by rate motifs found outside the degron, such as the KR motif [@moss_rate_2015; @dreher_arabidopsis_2006]. IAA-mediated repression of ARFs depends upon both an EAR domain, N-terminal of the degron, that recruits co-repressors of the TOPLESS (TPL) family, and a Phox and Bem1 (PB1) domain that enables interaction via hetero-oligomerization with ARF PB1 domains [@nanao_structural_2014; @kagale_ear_2011].

@@

###Sister Pair Diversification

Previous research has ~~pointed to~~ **found evidence of** both broad genetic redundancy and specificity within closely related pairs or groups of IAA proteins [@overvoorde_functional_2005]. For example, the iaa8-1, iaa9-1 double mutant and the iaa5-1, iaa6-1, iaa19-1 triple mutant have wild-type phenotypes [@overvoorde_functional_2005], yet the IAA6/19 sister pair has significant differences in expression patterns, protein abundances and functions, suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017].

In our analysis of the full IAA gene family, one member of most IAA sister pairs had high nonsynonymous diversity while the other sister was more conserved (Figure IAA phylogeny heat map). In an interspecific comparison between *A. thaliana* and *A. lyrata*, IAA19 is more conserved, while IAA6 showed regions of increased sequence divergence upstream of the degron and in the N-terminal half of the PB1 domain [@winkler_variation_2017]. Consistent with these results, our analysis of intraspecific natural variation revealed higher conservation for IAA19 (${\pi_N/\pi_S}$ = 0.55) compared to IAA6 (${\pi_N/\pi_S}$ = 2.3) (Figure IAA phylogeny heat map), and also detected high diversity within the same regions of IAA6 as seen in Winkler et al. (Supp Fig IAA6 diversity plot).

Within the Group B IAA genes, there are four sister pairs--IAA10/IAA11, IAA12/IAA13, IAA18/IAA26, and IAA20/IAA30 (Fig X) [@remington_contrasting_2004]. Interestingly, among these sister pairs only IAA12 and 13 have a noticeable difference in nonsynonymous diversity. The *Arabidopsis thaliana* EFP browser shows that IAA10 and 11 have almost identical expression patterns, with the exception of IAA10 being more highly expressed in the developing seed, and IAA11 being more highly expressed in root tissue during lateral root development. In our analysis, IAA10 and IAA11 both showed functional conservation (${\pi_N/\pi_S}$ of 0.80 and 0.67 respectively) in support of IAA10 and IAA11 playing redundant roles. Accessions and associated variants identified here provide a potentially expedited route to elucidating the extent of redundancy between family members such as IAA10 and 11.

###Identification of new Loss-of-Function/Gain-of-Function mutations
Accessions with functional variants can be potentially used to investigate the extent of redundancy/specialization within sister pairs. For example, it has been shown that IAA10 has a missense mutation in the EAR domain, which converts the second leucine in the LxLxL motif to a serine (L49S), a mutation likely to reduce the ability of the IAA to recruit the TOPLESS co-repressor [@lee_biological_2016]. This mutation was found in three accessions from the southern UK. A second potentially stronger mutation, represented in one accession from northeastern Spain is a ~~missense~~ **frameshift?** resulting in an early stop codon in the PB1 domain (need Supp. Fig?). This likely prevents IAA10 from interacting with an ARF. ~~These alleles would be strong candidates for loss-of-function mutants, allowing researchers to discover the function of IAA10. In particular, it would make possible a straightforward path for creating an iaa10/11 double mutant using CRISPR/Cas9 directed mutagenesis, allowing for examination of the extent of redundancy between these family members.~~ **I don't think we can say this. There are SALK lines in the 5' UTR and first intron of IAA10 and IAA11.**

Several examples of possible gain-of-function degron missense variants were found in IAA2, IAA3, IAA6, IAA7, IAA8, IAA13, IAA15, IAA18, IAA19, IAA27, IAA28, IAA29. Most degron variants were only found in one or two accessions (Supplemental Table or Map?). ~Interestingly, almost all of the IAAs found to have accessions with degron missense mutations have a closely related sister, which may indicate a redundant function within the sister-pairs.~ **But we know that iaa6 iaa19 iaa5 triple doesn't have a phenotype.** IAA6 has two potentially dominant missense mutations found within the degron (Supp Fig. alignment), one previously believed to lead to the shy1-1 phenotype [@kim_two_1996]. Interestingly, this variant is widely represented in 31 accessions across Europe (Supplemental Map Fig). This finding supports the proposition that iaa6[C78R] may not be the mutation causing the shy1-1 phenotype but rather represents natural variation in the degron sequence ([note from Mike Prigge in TAIR](https://www.arabidopsis.org/servlets/TairObject?type=notepad&id=11247)). These accessions could be used to confirm whether iaa6[C78R] is responsible for the SHY1-1D phenotype. Additionally, an accession was identified with the IAA6-V77M variant, also in the degron, that may be useful to further explore iaa6 gain-of-function phenotypes.

@@

Members of the IAA17/7/14/16 group have a second EAR motif (LxLxL) with no observed missense mutations in IAA7, 16, or 17 (Supp Fig IAA7/14/16/17 alignment). This observation supports the experimental evidence for IAA7 that the second EAR motif is functional: it can interact with TPR1 co-repressor and has a minor repressive role in auxin-related developmental phenotypes [lee_biological_2016]. For IAA14, we did observe two missense mutations that may be of interest to probe functionality of the IAA14 EAR motif: one in the central lysine (L41F) that is represented in only one accession, and the other in the second “x” (K40N) that is found in 12 accessions around Copenhagen (Supp Fig. IAA14 map).

@@

The Auxin signaling pathway utilizes the TOPLESS (TPL) and TOPLESS-related (TPR) family of Gro/TLE/TUP1 type co-repressor proteins to maintain auxin responsive genes in a transcriptionally-repressed state in the absence of auxin (cite). In *Arabidopsis thaliana* the TPL/TPR family includes *TPL* and four *TPR1-4*. The resulting proteins are comprised of three structural domains: an N-terminal TPL domain and two WD-40 domains. TPL/TPR proteins are recruited to the AUX/IAA proteins through interaction with the conserved Ethylene-responsive element binding factor-associated amphiphilic repression (EAR) domain (cite). Canonical EAR domains have the amino acid sequence LxLxL, as found in most AUX/IAAs (figure xA) [@overvoorde_functional_2005]. TPL/TPR co-repressors bind EAR domains via their C-terminal to LisH (CTLH) domains found near their N-termini (citations of pre-structure founding papers/reviews). Recent structural analyses of the TPL N-terminal domain have highlighted the precise interaction interface between TPL and AUX/IAA EAR domains, as well as the TPL-TPL dimerization and tetramerization motifs (Avarelilo et al 2017, Ke et al, 2015). The residues required for higher-order multimers of TPL tetramers have also been identified (Ma et al, 2017). Additional interactions with transcriptional regulation and chromatin modifying machinery are likely mediated by two tandem beta propeller domains of TPL/TPRs. These domain have not yet been crystallized, however numerous high resolution crystal structures of beta propeller domains allow for structural prediction of the TPL c-terminal beta propeller domains and the locations of likely interaction faces (Figure, X).

The TOPLESS co-repressor family generally exhibits a high level of sequence conservation at the amino acid sequence level across resequenced *Arabidopsis thaliana* accessions, with all ${\pi_N/\pi_S}$ values below 1 (Table X). The closely related TPL and TPR1 have the highest ${\pi_N/\pi_S}$ values (TPL-0.425, TPR1-0.739), suggesting that these these two related genes tolerate a higher degree of sequence and potentially functional diversity compared to TPR2-4. The N-terminal TPL domain of the TPL/TPR family is particularly conserved (Figure X). All non-synonymous polymorphisms observed in this region are either in the coils between helices or are highly conservative mutations within helices (i.e. Valine to Isoleucine), which would be predicted to exhibit little effect on TPL folding and function.

The high degree of conservation in the entire N-terminal domain underscores its importance. For example, the initial tpl-1 mutation (N176H) in the ninth helix is a dominant gain-of-function allele (Long et al, 2002, PMID12050130), which is capable of binding wild-type TPL protein and inducing protein aggregation (Ma, 2017, pmid28630893). It is therefore understandable that very little diversity is observed in the ninth alpha helix of TPL/TPRs as mutations in this domain could act in a dominant negative fashion, rendering any de novo mutation in this region embryo lethal. However, the entire N-terminal domain exhibits very low rates of nucleotide diversity, suggesting that each domain in this region is critical at the population level. Indeed, the crystal structure of TPL has revealed that the TPL domain is responsible for the multiple roles of TPL/TPR proteins: The LisH domain and CRA domain coordinate TPL dimerization and tetramerization respectively (26601214, 28698367), the CTLH domain is critical for interaction with DNA-binding transcription factors (26601214), and helix 9 within the CRA domain may be critical for the formation of higher-order TPL complexes (multimers of TPL tetramers, 28630893).

Compared to the TPL domain, both the proline rich and WD-40 domains exhibit a higher rate of missense mutations. We utilized the Phyre2.0 modeling software to predict the structure of the TPL/TPR WD-40 repeats in order to map non-synonymous mutations onto a visible protein chassis (Figure X). Here we can see both faces of each predicted beta-propeller structure of the WD-40 domains, which are connected by a short peptide linker. This model reveals amino acid conservation in the core and central residues of the WD-40 domains. ~~These observations could be used as the basis for forward mutagenesis strategies to test whether this region of TPL functions similarly to its homolog TLE, which binds its target genes EH1 and the C-terminal WRPW/Y motif (Hairy/Hes/Runx) in the center of the WD-40 domain (pmid 16762837).~~ **Can we say something more general here? There is not enough background to support this.**

Despite the presence of 5 TPL/TPR genes in the genome, which are considered to be redundant (pmid 16763149), very few SNPs are observed that are predicted to result in loss of function for any TPL/TPR gene. Only one frameshift variant  was observed, TPR4-665delG, in a single accession (Aiell-4 from Italy). This deletion occurs in codon 222, just after the TPL N-terminal domain, and generates an early stop 11 codons downstream from the deletion. It is unclear whether this N-terminal fragment would be expressed at any appreciable level, as it is possible that the generation of an early stop codon would trigger nonsense-mediated decay. However, if expressed, it would be fascinating to examine whether this naturally occurring N-terminal truncation retains activity as a co-repressor. Given the observation that the TPL N-terminal domain is sufficient to bind histones (28630893), and repress transcription via synthetic assays in yeast (24979769), it is possible that this represents a semi-functional TPL variant. There has been no evidence for prioritization of individual TPL/TPR family members in the auxin response pathway, but experiments modulating the gene copy number of TPR1 revealed the importance of maintaining the proper dosage of TPL/TPR in order to accurately trigger the salicylic acid innate immune response (20647385). These results, taken together with the absence of TPL/TPR loss-of-function alleles in the *Arabidopsis thaliana* accessions suggest that perhaps the entire family is being conserved at the nucleotide level in the population to meet a gene dosage requirement, with little opportunity for neo- or sub-functionalization.

@@

Auxin response is ultimately mediated by the auxin responsive transcription factors (ARFs). There are 23 ARFs in *Arabidopsis thaliana* that are divided into three phylogenetic classes. Class A ARFs (ARF5, ARF6, ARF7, ARF8 and ARF19) activate transcription. These ARFs have a glutamine-rich region in the middle of the protein that may mediate activation (Guilefoyle, 2007). It has recently been shown that the middle region of ARF5 interacts with the SWI/SNF chromatin remodeling ATPases BRAMA and SPLAYED, possibly to reduce nucleosome occupancy and allow for the recruitment of transcription machinery (Wu, 2016). Additionally, ARF7 interacts with Mediator subunits, directly tethering transcriptional activation machinery to its binding sites in the chromosome (Ito, 2016). Class B and C ARFs are historically categorized as repressor ARFs, though the mechanism through which they confer repression has not been identified. Their middle regions tend to be proline- and serine-rich (citation).

Canonical ARFs are comprised of three major domains. Recent crystallization of these domain have informed structure-function analysis of the ARFs (Boer, 2014; Nanao, 2014; Korasick, 2014). These domains are conserved throughout land plants. ARFs share an N-terminal B3 DNA binding domain. Flanking this DNA-binding domain is a dimerization domain, which folds up into a single “taco-shaped” domain to allow for dimerization between ARFs. There is an auxiliary domain that immediately follows and interacts with the dimerization domain. The middle region is the most variable between ARFs, as mentioned above, but is characterized by repetitive units of glutamine (class A), serine, or proline residues (classes B and C).
